Are mental properties supervenient on brain properties?

### 2010-03-13 manuscript/supervenience.tex
modified draft somewhat

%\input{/Users/joshyv/Research/misc/latex_paper.tex}
\documentclass{article}
\usepackage{fancyhdr}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{verbatim}
% \usepackage{ulem}
% \usepackage{color}

@@

\newcommand{\hL}{\widehat{L}}
\newcommand{\MeB}{\mM \overset{\varepsilon}{{\sim}}_F \mB}
\newcommand{\MnoteB}{\mM \overset{\varepsilon}{{\not\sim}}_F \mB}

@@

The ``mind-brain supervenience'' theorem suggests that all mental properties (e.g.\ consciousness, intelligence, personality, etc.) are derived from the physical properties of the brain. The validity of this theorem has been argued in philosphical terms for over 2,500 years, but it has not previously been approachable through experimental investigation. To address the question of whether the mind supervenes on the brain through empirical means, here we frame a supervenience hypothesis in rigorous mathematical terms and propose a modified version of supervenience (called $\varepsilon$-supervenience) that is amenable to scientific methods and statistical analysis. To elucidate this approach, we posit a thought experiment that illustrates how the probabilistic theory of pattern recognition can be used to make a one-sided determination of $\varepsilon$-supervenience. The physical property of the brain employed in this analysis is the graph describing brain connectivity (i.e., the \emph{connectome}), and $\varepsilon$-supervenience allows us to determine whether a particular mental property can be inferred from one's connectome to within any given misclassification rate $\varepsilon > 0$, regardless of the relationship between the two. In addition to the theoretical results, we show via simulation that given reasonable assumptions about class conditional probabilities and the amount of data available, the thought experiment can actually be conducted on a simple organism, \emph{Caenorhabditis elegans}, with currently available technology.

@@

@@

Unfortunately, it is in general impossible to determine whether $L_F(g^*)=0$ without knowing $F$. Therefore, we relax the above statistical supervenience to define $\varepsilon$-\textit{supervenience}:

@@

Now, generalizing the concept of a classifier $g$ to allow consideration of training data, consider $g_n:\mB \times (\mB \times \mM)^n \rightarrow \mM$ which takes as input an observed brain connectivity structure $b$ and $n$ training pairs $(b_1,m_1),\cdots,(b_n,m_n)$ and produces a classification $\widehat{m}=g_n(b;(b_1,m_1),\cdots,(b_n,m_n))$. Let $L_{F}(g_n) = E[P_{F}[g_n(B;(B_1,M_1),\cdots,(B_n,M_n)) \neq M|(B_1,M_1),\cdots,(B_n,M_n)]]$.

@@

Since the joint distribution $F=F_{BM}$ is unknown, the utility of Theorem \ref{thm1} requires that $g_n$ be a {\it universally consistent} classifier --- that is, $\lim_n L_{F}(g_n) = L_{F}(g^*)$ for all distributions $F=F_{BM}$. Unfortunately, the rate of convergence of $L_{F}(g_n)$ to $L_{F}(g^*)$ depends on the (unknown) distribution $F=F_{BM}$ \cite{DGL96}. Furthermore, arbitrarily slow convergence theorems regarding the rate of convergence of $L_{F}(g_n)$ to $L_{F}(g^*)$ demonstrate that there is no universal $n,n'$ which will guarantee that the test has power greater than any specified target $\beta > \alpha$ \cite{Devroye83}. For this reason, the test outlined above can provide only a one-sided conclusion: if we reject we can be $100(1-\alpha)$\% confident that $\MeB$ holds, but we can never be confident in its negation. Thus, without restrictions on $F_{BM}$, arbitrarily slow convergence theorems imply that our theorem of $\varepsilon$-supervenience does not satisfy Popper's {\it falsifiability} requirement \cite{Popper}. Given these caveats, consider the following thought experiment:

@@

Let the physical property under consideration be brain connectivity structure (``connectome''), so $b$ is a graph with vertices representing neurons (or neuroanatomical regions) and edges representing connections between neurons (or white matter tracts). Further let $\mB$, the observation space, be the collection of all graphs on a finite number of vertices, and let $|\mB|$ be countable. Now, imagine collecting very large amounts of very accurate independent and identically distributed brain-graph data and the associated mental property indicators. A $k_n$-nearest neighbor classifier using an isomorphism-matching Frobenius norm is universally consistent (see Appendix 1 for proof). Therefore, Theorem \ref{thm1} applies and the existence of a universally consistent classifier guarantees that eventually (in $n,n'$) we will be able to conclude $\MeB$ for this mental/brain property pair, if indeed $\varepsilon$-supervenience holds. This logic holds for directed graphs or multigraphs or hypergraphs with discrete edge weights and vertex attributes.

@@

Practical issues regarding actually conducting the above thought experiment include: (1) as stated, we must consider the space $\mB$ to be the quotient space of graphs mod graph isomorphism, unless the vertices are {\it labeled}; (2) a more informative and tractable distance on $\mB$ may be desired, as the $k_n$-nearest neighbor classifier under our Frobenius norm may have a rate of convergence so slow and a computational demand so high as to be impractical; and (3) collecting enough sufficiently accurate independent and identically distributed brain-graph data and the associated mental property indicators may be beyond current technological capabilities. Regardless, related experimental work includes collecting various types of brain graph data \cite{WhiteBrenner86, DenkHorstmann04, BriggmanDenk06} and various approaches to inference on brain graphs \cite{MackeBorst08, Mishchenko09, LuLichtman09}, suggesting feasibility of such an experiment in the near future (see Appendix 2 for a simulated example of a feasible experiment). Nevertheless, our thought experiment suggests that we can hope to determine that a given mental property under consideration $\varepsilon$-supervenes on a brain's connectivity structure.

@@

Assume first that all graphs are simple, on the same set of vertices, and that the graphs are labeled so that we know which vertex in one graph corresponds to which vertex in another. Then the Frobenius distance function $d(b_1,b_2)$ can be written in terms of the associated adjacency matrices $A_1$ and $A_2$: $d(b_1,b_2) = ||A_1-A_2||_F$. If the graphs are identical, then $d(b_1,b_2) = 0$, and if the graphs are different, then $d(b_1,b_2) \geq 1$. Since the space $\mB$ is finite, $n$ large enough guarantees that with probability approaching unity at least $k_n$ training samples coincide with each atom, so long as $k_n/n \rightarrow 0$. Then $k_n \rightarrow \infty$ guarantees that the nearest neighbor vote-winner for each atom will eventually coincide with Bayes' choice, yielding universal consistency.

@@

Several points of note: Isolated vertices are ignored in our equivalence relation; the class-conditional signal is entirely encompassed by the connectivity structure; the graph isomorphism problem is computationally hard \cite{ConroyLouis97,ZaslavskiyVert08}; and the argument employed here does not capture the concept of ``nearness implies likelihood of similar class''---we simply rely on atomic behavior.

@@

Simulations suggest that one may build a classifier, practically and with a manageable training sample size $n$, that demonstrates $\varepsilon$-supervenience with reasonable choices for $\varepsilon$ and $\alpha$ and a plausible joint distribution $F_{BM}$ (Figure \ref{fig1}). To generate the data, we let the class-conditional random variable $E_{ij} | M=m_0$ be distributed Poisson$(A_{ij}+\eta)$, where $A_{ij}$ is the number of chemical synapses between neuron $i$ and neuron $j$ according to \cite{VarshneyChklovskii09}, with noise parameter $0<\eta \ll 1$. The class-conditional random variable $E_{ij} | M=m_1$ is distributed Poisson$(A_{ij}+ k_{ij})$ for neurons $i,j \in \mD$, where $\mD$ is the set of edges deemed responsible for odor-evoked behavior according to \cite{ChalasaniBargmann07}, with signal parameter $k_{ij}$ uniformly sampled from $[-5,5]$. We consider $k_n$-nearest neighbor classification of labeled multigraphs (directed, with loops) on 279 vertices, under Frobenius norm. The $k_n$-nearest neighbor classifier used here satisfies $k_n \rightarrow \infty$ as $n \rightarrow \infty$ and $k_n/n \rightarrow 0$ as $n \rightarrow \infty$, ensuring universal consistency. (Better classifiers can be constructed for the joint distribution $F_{BM}$ used here; however, we demand universal consistency.)

@@

%\subparagraph*{References}
\bibliography{biblist}
\addcontentsline{toc}{section}{References}
%\bibliography{Science}
\bibliographystyle{unsrt}
% \bibliographystyle{nature}

\begin{figure}[h!]
\centering \includegraphics[width=.9\linewidth]{Lhatplot}
\caption{C.~elegans graph classification simulation results. $\hL^{1000}_{F}(g_n)$ is plotted as a function of class-conditional training sample size $n_j$, suggesting that for $\varepsilon=0.1$ we can determine that $\MeB$ holds with $99\%$ confidence with just a few hundred training samples generated from $F_{BM}$. Each dot depicts an estimate for $L_{F}(g_n)$; standard errors are $(L_{F}(g_n)(1-L_{F}(g_n))/1000)^{1/2}$. E.g., $n_j = 180$ ; $k_n = 53$ ; $\hL^{1000}_{F}(g_n) = 0.057$; standard error less than 0.01. We reject $H_0: L_{F}(g^*) \geq 0.10$ at $\alpha=0.01$. $L_{F}(g^*) \approx 0$ for this simulation.
}
\label{fig1}
\end{figure}

>>>>

%\input{/Users/joshyv/Research/misc/latex_paper.tex}
\documentclass{article}
\usepackage{fancyhdr}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{color}
% \usepackage{ulem}
% \usepackage{color}

@@

\newcommand{\hL}{\widehat{L}}
\newcommand{\MeB}{\mM \overset{\varepsilon}{{\sim}}_F \mB}
\newcommand{\MnoteB}{\mM \overset{\varepsilon}{{\not\sim}}_F \mB}
\providecommand{\tr}[1]{\textcolor{red}{#1}}

@@

The ``mind-brain supervenience'' theorem suggests that all mental properties (e.g.\ consciousness, intelligence, personality, etc.) are derived from the physical properties of the brain. The validity of this theorem has been argued in philosophical terms for over 2,500 years. \tr{The alternative hypotheses, including various non-physical causes of mental properties, seems rather implausible to many.  That said, testing these hypotheses } has not previously been approachable through experimental investigation. To address the question of whether the mind supervenes on the brain through empirical means, here we frame a supervenience hypothesis in rigorous mathematical terms and propose a modified version of supervenience (called $\varepsilon$-supervenience) that is amenable to scientific methods and statistical analysis. To elucidate this approach, we posit a thought experiment that illustrates how the probabilistic theory of pattern recognition can be used to make a one-sided determination of $\varepsilon$-supervenience. The physical property of the brain employed in this analysis is the graph describing brain connectivity (i.e., the \emph{connectome}), and $\varepsilon$-supervenience allows us to determine whether a particular mental property can be inferred from one's connectome to within any given misclassification rate $\varepsilon > 0$, regardless of the relationship between the two. In addition to the theoretical results, we show via simulation that given reasonable assumptions about class conditional probabilities and the amount of data available, the thought experiment can actually be conducted on a simple organism, \emph{Caenorhabditis elegans}, with currently available technology.

@@

\tr{The primary contributions of this work are as follows.  First, a notion of supervenience emendable to empirical investigation is formally introduced.  This renders the mind-brain dualism debate a hypothesis, rather than an assumption.  Second, the limits of hypothesis testing to unravel supervenience hypotheses are provided.  Third, we posit a very general model of brains and their associated mental properties that admits a universally consistent classifier; that is, we prove that assuming the brain can be accurately represented by this model, and there is sufficient data, this algorithm is guaranteed to obtain the best possible results.  Fourth, simulations show this algorithm converges relatively rapidly.}

@@

\tr{(Note that this definition does not imply a one-to-one mapping.)}  \tr{To allow for the possibility of only \emph{partial} supervenience,} \footnote{The following was stricken from the record, as I believed it to be a non-sequitor, but i partially replace it below: Unfortunately, it is in general impossible to determine whether $L_F(g^*)=0$ without knowing $F$. Therefore,} we relax the above statistical supervenience to define $\varepsilon$-\textit{supervenience}:

@@

\tr{Unfortunately, in general $F$ is unknown, but can be estimated from the data. Therefore,} generalizing the concept of a classifier $g$ to allow consideration of training data, consider $g_n:\mB \times (\mB \times \mM)^n \rightarrow \mM$ which takes as input an observed brain connectivity structure $b$ and $n$ training pairs $\tr{\vec{d}_n=}(b_1,m_1),\cdots,(b_n,m_n)$ and produces a classification $\widehat{m}=g_n(b;\tr{\vec{d}_n})$. Let $L_{F}(g_n) = E[P_{F}[g_n(B;\tr{\vec{D}_n}) \neq M|\tr{\vec{D}_n}]]$.

@@

Since the joint distribution $F=F_{BM}$ is unknown, the utility of Theorem \ref{thm1} requires that $g_n$ be a {\it universally consistent} classifier --- that is, $\lim_n L_{F}(g_n) = L_{F}(g^*)$ for all distributions $F=F_{BM}$. Unfortunately, the rate of convergence of $L_{F}(g_n)$ to $L_{F}(g^*)$ depends on the (unknown) distribution $F=F_{BM}$ \cite{DGL96}. Furthermore, arbitrarily slow convergence theorems regarding the rate of convergence of $L_{F}(g_n)$ to $L_{F}(g^*)$ demonstrate that there is no universal $n,n'$ which will guarantee that the test has power greater than any specified target $\beta > \alpha$ \cite{Devroye83}. For this reason, the test outlined above can provide only a one-sided conclusion: if we reject we can be $100(1-\alpha)$\% confident that $\MeB$ holds, but we can never be confident in its negation. Thus, without restrictions on $F_{BM}$, arbitrarily slow convergence theorems imply that our theorem of $\varepsilon$-supervenience does not \tr{strictly} satisfy Popper's {\it falsifiability} requirement \cite{Popper}.\footnote{\tr{The strictly must be there because it is possible for the null to be rejected, given the data has a particular distribution, it is simply not necessarily rejectable.  This is a subtle point that was lost on me until very recently.}} Given these caveats, consider the following thought experiment:

@@

Let the physical property under consideration be brain connectivity structure (``connectome''), so $b$ is a graph \tr{(or, network)} with vertices representing neurons (or neuroanatomical regions) and edges representing connections between neurons (or white matter tracts). Further let $\mB$, the observation space, be the collection of all graphs on a finite number of vertices, and let $|\mB|$ be countable. Now, imagine collecting very large amounts of very accurate independent and identically distributed brain-graph data and the associated mental property indicators. A $k_n$-nearest neighbor classifier using an isomorphism-matching Frobenius norm is universally consistent (see Appendix 1 for proof). Therefore, Theorem \ref{thm1} applies and the existence of a universally consistent classifier guarantees that eventually (in $n,n'$) we will be able to conclude $\MeB$ for this mental/brain property pair, if indeed $\varepsilon$-supervenience holds. This logic holds for directed graphs or multigraphs or hypergraphs with discrete edge weights and vertex attributes.

@@

Practical issues regarding actually conducting the above thought experiment include: (1) as stated, we must consider the space $\mB$ to be the quotient space of graphs mod graph isomorphism, unless the vertices are {\it labeled}; (2) a more informative and tractable distance on $\mB$ may be desired, as the $k_n$-nearest neighbor classifier under our Frobenius norm may have a rate of convergence so slow and a computational demand so high as to be impractical; and (3) collecting enough sufficiently accurate independent and identically distributed brain-graph data and the associated mental property indicators may be beyond current technological capabilities. Regardless, related experimental work includes collecting various types of brain graph data \cite{WhiteBrenner86, DenkHorstmann04, BriggmanDenk06} and various approaches to inference on brain graphs \cite{MackeBorst08, Mishchenko09, LuLichtman09}, suggesting feasibility of such an experiment in the near future (see Appendix 2 for a simulated example of a feasible experiment). Nevertheless, our thought experiment suggests that we can hope to determine that a given mental property under consideration $\varepsilon$-supervenes on a brain's connectivity structure. \tr{This thought experiment, together with (i) the formal definition of $\varepsilon$-supervenience, (ii) the brain-graph model, and (iii) the universal consistency proof on graphs, is the first \emph{proof} (to our knowledge) that ``hope'' of empirically investigating supervenience is possible.}

@@

Assume first that all graphs are simple \tr{(meaning undirected with no loops and binary binary edges)}, on the same set of vertices, and that the graphs are labeled so that we know which vertex in one graph corresponds to which vertex in another. Then the Frobenius distance function $d(b_1,b_2)$ can be written in terms of the associated adjacency matrices $A_1$ and $A_2$: $d(b_1,b_2) = ||A_1-A_2||_F$. If the graphs are identical, then $d(b_1,b_2) = 0$, and if the graphs are different, then $d(b_1,b_2) \geq 1$. Since the space $\mB$ is finite, $n$ large enough guarantees that with probability approaching unity at least $k_n$ training samples coincide with each atom, so long as $k_n/n \rightarrow 0$. Then $k_n \rightarrow \infty$ guarantees that the nearest neighbor vote-winner for each atom will eventually coincide with Bayes' choice, yielding universal consistency.

@@

Several points of note: isolated vertices are ignored in our equivalence relation; the class-conditional signal is entirely encompassed by the connectivity structure; the graph isomorphism problem is computationally hard \cite{ConroyLouis97,ZaslavskiyVert08}; and the argument employed here does not capture the concept of ``nearness implies likelihood of similar class''---we simply rely on atomic behavior.

@@

Simulations suggest that one may build a classifier, practically and with a manageable training sample size $n$, that demonstrates $\varepsilon$-supervenience with reasonable choices for $\varepsilon$ and $\alpha$ and a plausible joint distribution $F_{BM}$ (Figure \ref{fig1}). To generate the data, we let the class-conditional random variable $E_{ij} | M=m_0$ be distributed Poisson$(A_{ij}+\eta)$, where $A_{ij}$ is the number of chemical synapses between neuron $i$ and neuron $j$ according to \cite{VarshneyChklovskii09}, with noise parameter $0<\eta \ll 1$. The class-conditional random variable $E_{ij} | M=m_1$ is distributed Poisson$(A_{ij}+ \tr{z_{ij}})$ for neurons $i,j \in \mD$, where $\mD$ is the set of edges deemed responsible for odor-evoked behavior according to \cite{ChalasaniBargmann07}, with signal parameter $\tr{z_{ij}}$ uniformly sampled from $[-5,5]$. We consider $k_n$-nearest neighbor classification of labeled multigraphs (directed, with loops) on 279 vertices, under Frobenius norm. The $k_n$-nearest neighbor classifier used here satisfies $k_n \rightarrow \infty$ as $n \rightarrow \infty$ and $k_n/n \rightarrow 0$ as $n \rightarrow \infty$, ensuring universal consistency. (Better classifiers can be constructed for the joint distribution $F_{BM}$ used here; however, we demand universal consistency.)

@@

%\subparagraph*{References}
\bibliography{biblist}
\addcontentsline{toc}{section}{References}
%\bibliography{Science}
\bibliographystyle{ieeetr}
% \bibliographystyle{nature}

\begin{figure}[h!]
\centering \includegraphics[width=.9\linewidth]{Lhatplot}
\caption{C.~elegans graph classification simulation results. $\hL^{1000}_{F}(g_n)$ is plotted as a function of class-conditional training sample size $n_j$, suggesting that for $\varepsilon=0.1$ we can determine that $\MeB$ holds with $99\%$ confidence with just a few hundred training samples generated from $F_{BM}$. Each dot depicts an estimate for $L_{F}(g_n)$; standard errors are $(L_{F}(g_n)(1-L_{F}(g_n))/1000)^{1/2}$\tr{; e}.g., $n_j = 180$ ; $k_n = 53$ ; $\hL^{1000}_{F}(g_n) = 0.057$; standard error less than 0.01. We reject $H_0: L_{F}(g^*) \geq 0.10$ at $\alpha=0.01$. $L_{F}(g^*) \approx 0$ for this simulation.
}
\label{fig1}
\end{figure}
